Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2021-01-26 manuscript_Huang.tex
revision almost done!

For all four situations, we investigate the SCMs along with their dual entropy characteristics in detail and summarize the obtained results in Tab.~\ref{tableLog}. We note that several types of chaos--chaos transitions will be discussed in the following: the band merging crisis corresponds to intermittency, the inner crisis to some more subtle chaos-chaos transition and the outer crisis to fully chaotic dynamics.

\begin{table}[htb]
    {\begin{tabular}{l  l  l  l  l  l  l  l}
    \hline
    Regime & $r$      & $\mathcal{H}_O$ & $\mathcal{C}_O$ & $\mathcal{H}_W$ & $\mathcal{C}_W$ & $\mathcal{H}_E$   & $\mathcal{C}_E$  \\
    \hline
    period-3      & 3.83 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \hline
    band merging    & 3.679 & 0.978 & 0.043 & 0.595 & 0.577 & 0.211 & 0.205  \\
    \hline
    laminar state    & 3.791 & 0.986 & 0.029 & 0.633 & 0.609 & 0.280 & 0.269 \\
    \hline
    outer crisis  & 4.0 & 0.994 & 0.014 & 0.677 & 0.638 & 0.359 & 0.339 \\
    \hline
    \end{tabular}}
   \caption{Values of the three considered SCMs along with their adjoint entropy measures for four different values of the control parameter $r$ of the logistic map (embedding parameters: $D = \tau = 6$).   \label{tableLog}}    
\end{table}

Due to our choice of $\tau$ and $D$, all SCMs take values of zero for the period-3 series because the temporal distance between each pair of components of the embedding vector covers a multiple of two full periods (cf.~Fig.~\ref{fig:embed}). Other choices of $D$ and $\tau$ would commonly yield non-zero values for the period-3 regime. For the three other cases of band merging, laminar state and outer crisis, the values of all SCMs significantly differ from zero. The pattern frequency based SCM $\mathcal{C}_O$ shows smaller values than the other two measures for all regimes, which implies a positioning in the lower right part of the complexity--entropy plane in Fig.~\ref{fig:CElogistic}(a). However, for the transition matrix based SCMs, both the estimated entropies $\mathcal{H}_W$, $\mathcal{H}_E$ and associated complexity measures $\mathcal{C}_W$, $\mathcal{C}_E$ show results that are more consistent with the expectations when considering the respective level of chaoticity of the system in those three regimes (i.e., a rising Lyapunov exponent with increasing $r$ in the respective chaotic regimes). In particular, the estimated SCMs take their largest values for $r = 4$ in the outer crisis regime, and the SCMs are the smallest when $r = 3.679$ in the band merging case, while the laminar regime displays intermediate SCM values.

\subsection{Complexity--entropy planes} \label{sec:plane}

In order to put our previous results into a broader context, we next show the complexity--entropy planes obtained when varying the control parameter $r$ in the logistic map. This simple nonlinear system has already been widely discussed in the framework of SCMs \cite{RossoPRE2007,MartinPLA2003}, but only for pattern frequency based complexity measures. For each value of $r$, we construct an OPTN from a time series of length $L = 10 ^ 6$ with the embedding parameters $D = \tau \in \{2,\ldots,7\}$.

As already stated above, when only one unique ordinal pattern is identified in a given time series, all SCMs take a value of $0$, which is reasonable since only self-loops are observed. This however happens only if the embedding parameters are chosen such that they exactly coincide with the periodicity of a certain periodic window (see Fig.~\ref{fig:embed}(g)). Another practical aspect concerns the data requirements for computing entropies and complexity measures. For a reliable estimation of SCMs, a pragmatic condition on the length $L$ of the time series to be sufficiently larger than $D!$ \cite{rossoPRL2007,kowalskiPhyD2007} has been satisfied in our setting for $D \leq 7$. In the particular example of the logistic map, one may easily generate much longer realizations, which however do not improve the obtained entropy and SCM estimates significantly (as we have tested for $L \leq 5 \times 10^6$, not shown).

@@

Figure~\ref{fig:CElogistic} shows the complexity--entropy planes for embedding parameters $D = \tau$ varied between $4$ and $6$. We emphasize that the results are qualitatively similar when $D$ and $\tau$ are varied in the range $\{3,\ldots, 7\}$. Furthermore, it is notable that the values of $\mathcal{C}_{max}$ depend on the embedding dimension $D$ since $D$ determines the number of patterns and pattern transitions which are considered in the definition of SCMs. By contrast, $\mathcal{C}_{min}$ shows less dependence on $D$.

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_LogisticHC.pdf}
\caption{\small{(Color online) Three different complexity--entropy planes (based on the Jensen-Shannon divergence) illustrating the behavior of the three SCMs for example series of the logistic map with varying control parameter $r$: (a, d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b, e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c, f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The dashed lines correspond to the maximum $\mathcal{C}_{max}$ and minimum complexity $\mathcal{C}_{min}$ values in dependence on the respective entropy. Panels (a-c) show the results for embedding dimensions $D = 4,5$ and (d-f) for $D = 6$. In panels (a, d), the bottom-right rectangles of dotted red lines indicate the regions for the traditional pattern frequency based SCM analysis where the differentiation between stochastic and deterministic--chaotic dynamics is ambiguous. }  \label{fig:CElogistic}}
\end{figure*}

@@

For our new pattern transition based SCMs, we find that in the complexity--entropy planes $\mathcal{H}_W \times \mathcal{C}_W$ and $\mathcal{H}_E \times \mathcal{C}_E$, the results for the logistic map again closely align with the theoretical upper bounds $\mathcal{C}_{max}$. When increasing the embedding dimension, the estimated values approach the theoretical $\mathcal{C}_{max}$ curve more and more closely and essentially cover exclusively the left uprising branch of the curve for the SCM based solely on transition frequencies (Figs.~\ref{fig:CElogistic}(b, e) and \ref{fig:CElogistic}(c, f)), which distinctively differs from the original permutation entropy based SCM. Specifically, the definition of $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ takes information on both, pattern frequencies and transition frequencies into account, which results in the obtained SCM values aligning even more closely with $\mathcal{C}_{max}$ already at lower embedding dimensions (Fig.~\ref{fig:CElogistic}(c, f)). In our opinion, the most interesting result is, however, that we do not find any values in the lower right part of the plane, which for the classical SCM based on permutation entropy would present the challenging region of possible ambiguity between deterministic-chaotic and stochastic dynamics. {\color{red}On the other hand, there are many points in the lower left part of the plane showing small values of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$. } We therefore suggest that it is the deterministic transition behavior among patterns that allows an improved distinction from random noise.

In order to further analyze the capability of our modified complexity--entropy planes to distinguish chaotic from stochastic dynamics, we compare the results for the logistic map with those obtained for fractional Brownian motion (fBm), i.e., stochastic processes with long-range temporal correlations. Specifically, for an fBm process the long-range correlations of the process are uniquely characterized by the Hurst exponent $H$ when positively correlated (persistence) for $1/2 < H < 1$, while being suppressed (anti-persistence) for $0< H < 1/2$. $H=1/2$ corresponds to the classical Brownian motion. Similar to the previous case of the logistic map with varying control parameter $r$, we generate time series for varying $H \in (0, 1)$ with a step size of $\Delta H=0.05$ and present the resulting complexity--entropy planes in Fig.~\ref{fig:CEfBm}. Moreover, since fBm series in the persistent regime are nonstationary, we further transform all fBm realizations into stationary time series by employing first-order difference filtering, i.e., by considering the increments $x_{i+1}- x_i$. The transformed series are commonly referred to as fractional Gaussian noise (fGn), and the associated complexity--entropy planes are shown in Fig.~\ref{fig:CEfGn}. Notably, fGn retains the long-range correlations and Gaussian probability density function (PDF) of the underlying fBm process. For a discussion on practical concerns regarding time delay embedding of fBm and fGn processes, we refer to Ref.~\cite{Zou2015}. 
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_fBm.pdf}
\caption{\small{(Color online) Same as in Fig.~\ref{fig:CElogistic}, but for fractional Brownian motion with a Hurst exponent $H \in (0, 1)$ varying with a step size of $\Delta H=0.05$. Note that the normalized entropy is generally decreasing with increasing $H$ (red arrow).}  \label{fig:CEfBm}}
\end{figure*}

@@

Comparing the results for both types of stochastic processes with those for the logistic map, a few important observations are made. First of all, the traditional pattern frequency based SCM values as originally reported by Rosso and co-workers \cite{RossoPRE2007,rossoPRL2007} are well reproduced as expected (Figs.~\ref{fig:CEfBm}(a, d) and \ref{fig:CEfGn}(a, d)). Second, all complexity--entropy pairs of fBm and fGn clearly differ from the theoretical maximum complexity values, which does not change with increasing embedding dimension $D$. Especially for the new OPTN based SCMs, we find a partial overlap between the entropy values of fBm and the logistic map, while the associated SCM values are however clearly distinct except for the cases with the lowest entropy, which correspond to the most nonstationary situations ($H$ close to 1). {\color{red}Taking into account the effects of self loops in the transition matrix, the results of Fig. \ref{fig:CEfBm} do not change significantly although the SCM values extend slightly in the left half of the complexity--entropy plane as $H$ is increased, which are shown in Fig. S1 of Supplementary Material (SM). Note that the discriminatory power of } Accordingly, we suggest that combining information on the position in the complexity--entropy plane and on the stationarity could allow a clear distinction between chaotic and stochastic dynamics even in those extreme situations. Notably, for the stationary fGn, all entropy measures are confined to very high entropy and low complexity values -- a range where no combinations have been found for the new transition frequency based SCMs for the logistic map. {\color{red}In the case of fGn, almost the same complexity--entropy plane has been obtained when self loops are included, which has been shown in Fig. S3 of SM. }

@@

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{logisticEntropy.pdf}
\caption{\small{Behavior of the different SCMs and associated entropy characteristics for the logistic map in dependence on the control parameter $r$. (a) Bifurcation diagram, (b) Lyapunov exponent, (c) pattern frequency based (permutation) entropy $\mathcal{H}_O$ and (d) associated complexity measure $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{H}_W$ based on the globally normalized transition matrix $\mathbf{W}$ and (f) the corresponding complexity measure $\mathcal{C}_W$, and (g, h) entropy $\mathcal{H}_E$ and complexity measure $\mathcal{C}_E$ based on the node-wise out-link normalized transition matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background shading. Vertical dotted lines indicate the cases summarized in Tab.~\ref{tableLog}. } \label{fig:bifurcation}}
\end{figure*}

@@

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{rosslerEntropy.pdf}
\caption{\small{Same as Fig. \ref{fig:bifurcation}, but for Poincar\'e sections of the R\"ossler system while varying the control parameter $b$ (see text for details). Panel (a) presents a bifurcation diagram based on the $x$ components of all points in the Poincar\'e sections. Panel (b) shows the largest Lyapunov exponent of the system in dependence of $b$.} \label{fig:bifurRossler}}
\end{figure*}

@@

Some open problems also common to other nonlinear time series analysis methods remain to be further addressed, for example, the dependence of reliable estimates of SCMs on the available time series length. {\color{red}For reliable estimations of traditional entropy $\mathcal{H}_O$ and SCM $\mathcal{C}_O$, the length $N > 5 D!$ has been largely suggested \cite{BandtPRL2002,AmigoPTRSA2014}. However, even longer length is required for the newly proposed SCMs based on transition frequencies because the transition matrix has the dimension $D!\times (D! - 1)$ if self loops are excluded. This should not be a big problem when working with time series from numerical models provided with the rapid increased computing power nowadays. } For time-discrete dynamical systems like chaotic maps, one may easily use longer data series, which however does not seem to improve the estimates very much due to the accumulation of effects due to a limited numerical precision (Fig.~\ref{fig:transient}). {\color{red} The dependence of the new SCMs on the length $N$ becomes crucial for experimental time series studies.} For continuous systems, together with the embedding parameters, the sampling frequency is another important characteristic which needs to be further explored. Moreover, the generalization of our approach from univariate to bivariate time series analysis shall be further discussed. We noticed that this aspect has remained a largely untouched subject for statistical complexity analysis so far, while it presents an interesting task to show the dependence of complexity on interaction. {\color{red} We may integrate some of the traditional network topological statistical measures of higher order into the SCMs, for instance, clustering coefficients or path lengths, which have been widely discussed to characterize the dynamical properties of the system generating the time series at hand \cite{ZouPR2018}.} We outline further work on the aforementioned topics as subjects of future research.

\section*{Acknowledgements}
Parts of this work have been financially supported by the National Natural Science Foundation of China (grant no. 11872182,11835003, and 11875132), the Natural Science Foundation of Shanghai (grant no. 18ZR1411800). YZ further acknowledges support by the Shanghai Municipal Science and Technology Major Project (grant no. 2018SHZDZX01) and ZJ Lab. The authors thank J. B. Borges and O. A. Rosso for sharing algorithms for computing $\mathcal{C}_{min}$ and $\mathcal{C}_{max}$.

>>>>

For all four situations, we investigate the SCMs along with their dual entropy characteristics in detail and summarize the obtained results in Tab.~\ref{tableLog}. We note that several types of chaos--chaos transitions will be discussed in the following: the band merging crisis corresponds to intermittency, the inner crisis to some more subtle chaos-chaos transition and the outer crisis to fully chaotic dynamics. 
\begin{table}[htb]
    {\begin{tabular}{l  l  l  l  l  l  l  l}
    \hline
    Regime & $r$      & $\mathcal{H}_O$ & $\mathcal{C}_O$ & $\mathcal{H}_W$ & $\mathcal{C}_W$ & $\mathcal{H}_E$   & $\mathcal{C}_E$  \\
    \hline
    period-3      & 3.83 & 0 & 0 & 0 & 0 & 0 & 0 \\
    \hline
    band merging    & 3.679 & 0.978 & 0.043 & 0.639 & 0.582 & 0.302 & 0.277  \\
    \hline
    laminar state    & 3.791 & 0.989 & 0.026 & 0.671 & 0.618 & 0.354 & 0.328 \\
    \hline
    outer crisis  & 4.0 & 1.0 & 0 & 0.761 & 0.662 & 0.522 & 0.458 \\
    \hline
    \end{tabular}}
   \caption{Values of the three considered SCMs along with their adjoint entropy measures for four different values of the control parameter $r$ of the logistic map (embedding parameters: $D = \tau = 6$).   \label{tableLog}}    
\end{table}

Due to our choice of $\tau$ and $D$, all SCMs take values of zero for the period-3 series because the temporal distance between each pair of components of the embedding vector covers a multiple of two full periods (cf.~Fig.~\ref{fig:embed}). Other choices of $D$ and $\tau$ would commonly yield non-zero values for the period-3 regime. For the three other cases of band merging, laminar state and outer crisis, the values of all SCMs significantly differ from zero. The pattern frequency based SCM $\mathcal{C}_O$ shows zero value and $\mathcal{H}_{O} = 1$, which implies a positioning in the lower right corner of the complexity--entropy plane in Fig.~\ref{fig:CElogistic}(a). However, for the transition matrix based SCMs, both the estimated entropies $\mathcal{H}_W$, $\mathcal{H}_E$ and associated complexity measures $\mathcal{C}_W$, $\mathcal{C}_E$ show results that are more consistent with the expectations when considering the respective level of chaoticity of the system in those three regimes (i.e., a rising Lyapunov exponent with increasing $r$ in the respective chaotic regimes). In particular, the estimated SCMs take their largest values for $r = 4$ in the outer crisis regime, and the SCMs are the smallest when $r = 3.679$ in the band merging case, while the laminar regime displays intermediate SCM values.

\subsection{Complexity--entropy planes} \label{sec:plane}
\subsubsection{Logistic map} 
In order to put our previous results into a broader context, we next show the complexity--entropy planes obtained when varying the control parameter $r$ in the logistic map. This simple nonlinear system has already been widely discussed in the framework of SCMs \cite{RossoPRE2007,MartinPLA2003}, but only for pattern frequency based complexity measures. For each value of $r$, we construct an OPTN from a time series of length $L = 10 ^ 6$ with the embedding parameters $D = \tau \in \{2,\ldots,7\}$.

As already stated above, when only one unique ordinal pattern is identified in a given time series, all SCMs take a value of $0$, which is reasonable since only self-loops are observed. This however happens only if the embedding parameters are chosen such that they exactly coincide with the periodicity of a certain periodic window (see Fig.~\ref{fig:embed}(g)). Another practical aspect concerns the data requirements for computing entropies and complexity measures. For a reliable estimation of SCMs, a pragmatic condition on the length $L$ of the time series to be sufficiently larger than $D!$ \cite{rossoPRL2007,kowalskiPhyD2007} has been satisfied in our setting for $D \leq 7$. In the particular example of the logistic map, one may easily generate much longer realizations, which however {\color{red} is subjected to finite size effects in experimental time series analysis}.

@@

Figure~\ref{fig:CElogistic} shows the complexity--entropy planes for embedding parameters $D = \tau$ varied between $4$ and $6$. We emphasize that the results are qualitatively similar when $D$ and $\tau$ are varied in the range $\{3,\ldots, 7\}$. Furthermore, it is notable that the values of $\mathcal{C}_{max}$ depend on the embedding dimension $D$ since $D$ determines the number of patterns and pattern transitions which are considered in the definition of SCMs. {\color{red}The curves of $\mathcal{C}_{max}$ for different embeddings have been further annotated in Fig. \ref{fig:CElogistic}. In contrast, the $\mathcal{C}_{min}$ values are close to each other, showing less dependence on $D$, and therefore they are not annotated. }
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_LogisticHC.pdf}
\caption{\small{(Color online) Three different complexity--entropy planes (based on the Jensen-Shannon divergence) illustrating the behavior of the three SCMs for example series of the logistic map with varying control parameter $r$: (a, d) $\mathcal{C}_O$ versus $\mathcal{H}_O$, (b, e) $\mathcal{C}_{W}$ versus $\mathcal{H}_{W}$, and (c, f) $\mathcal{C}_{E}$ versus $\mathcal{H}_{E}$. The {\color{red} dotted lines correspond to $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ in dependence on the respective entropy for $D=4$, and the dashed lines are $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ for $D=5$. Panels (a-c) show the results for embedding dimensions $D = 4,5$ with annotations for the respective $\mathcal{C}_{max}$ only} and (d-f) for $D = 6$. In panels (a, d), the bottom-right rectangles of dotted red lines indicate the regions for the traditional pattern frequency based SCM analysis where the differentiation between stochastic and deterministic--chaotic dynamics is ambiguous. However, the lower left rectangles of (b, c, e, f) correspond to regions of lower complexities. }  \label{fig:CElogistic}}
\end{figure*}

@@

For our new pattern transition based SCMs, we find that in the complexity--entropy planes $\mathcal{H}_W \times \mathcal{C}_W$ and $\mathcal{H}_E \times \mathcal{C}_E$, the results for the logistic map again closely align with the theoretical upper bounds $\mathcal{C}_{max}$. When increasing the embedding dimension, the estimated values approach the theoretical $\mathcal{C}_{max}$ curve more and more closely and essentially cover exclusively the left uprising branch of the curve for the SCM based solely on transition frequencies (Figs.~\ref{fig:CElogistic}(b, e) and \ref{fig:CElogistic}(c, f)), which distinctively differs from the original permutation entropy based SCM. Specifically, the definition of $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ takes information on both, pattern frequencies and transition frequencies into account, which results in the obtained SCM values aligning even more closely with $\mathcal{C}_{max}$ already at lower embedding dimensions (Fig.~\ref{fig:CElogistic}(c, f)). In our opinion, the most interesting result is, however, that we do not find any values in the lower right part of the plane, which for the classical SCM based on permutation entropy would present the challenging region of possible ambiguity between deterministic-chaotic and stochastic dynamics.

{\color{red}On the other hand, there are many values in the lower left corner of the plane showing small values of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ (Fig. \ref{fig:CElogistic}(b, c, e, f)). This region corresponds to regular orbits of low complexities. Along with the choice of embedding parameters, the length of time series and long term accumulations of the computation precisions play important roles in the lower left region of the complexity planes, which will be further illustrated by Figs. \ref{fig:bifurcation} and \ref{fig:transient}. Out of the left and right corners, complexities of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ show consistent trends along with the degree of chaoticity as will be explained later in Fig. \ref{fig:bifurcation}. } In general, we therefore suggest that it is the deterministic transition behavior among patterns that allows an improved distinction from random noise.

{\color{red}In addition, we observe a linear correlation between $\mathcal{C}_W$ and $\mathcal{H}_{W}$ (Fig. \ref{fig:CElogistic}(e)), respectively, $\mathcal{C}_{E}$ and $\mathcal{H}_{E}$ (Fig. \ref{fig:CElogistic}(f)). However, we note that this linear correlation appears only for large embedding $D = 6$. This linear correlation coincides with the up-rising brach of the maximal complexity curve $\mathcal{C}_{max}$, which is expected by the theory \cite{SMCbook2010}. }

{\color{red} @ Reik: I'm not sure whether to include the following discussion in the main text as suggested by the referee. The discriminatory power of different SCMs have been further compared in Fig. \ref{fig:CHlogCmax}, which characterize the distances to the maximal complexity $\mathcal{C}_{max}$. In particular, for a given $\mathcal{H}$ we compute a normalized complexity $\mathcal{C}^{N} = (\mathcal{C} - \mathcal{C}_{min}) / (\mathcal{C}_{max} - \mathcal{C}_{min})$. In turn, $\mathcal{C}^{N}$ is bounded in the interval of $[0, 1]$ and the alignment along $\mathcal{C}_{max}$ is reflected by large values of $\mathcal{C}^{N}$. The results are shown in Fig. \ref{fig:CHlogCmax}. The dots larger than 1 are numerical artefacts Fig. \ref{fig:CHlogCmax}(c) which are due to  improper choice of embeddings when the Logistic map shows high periodic iterations.    
\begin{figure*}
        \centering
        \includegraphics[width=2\columnwidth]{CompEntropyCNormalized_Logistic.pdf}
\caption{(Color online) Three SCMs which are normalized by $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$  at a given $\mathcal{H}$, for the logistic map.  \label{fig:CHlogCmax}}
\end{figure*}
}

\subsubsection{Fractional Brownian motion and fractional Gaussian noise}
In order to further analyze the capability of our modified complexity--entropy planes to distinguish chaotic from stochastic dynamics, we compare the results for the logistic map with those obtained for fractional Brownian motion (fBm), i.e., stochastic processes with long-range temporal correlations. Specifically, for an fBm process the long-range correlations of the process are uniquely characterized by the Hurst exponent $H$ when positively correlated (persistence) for $1/2 < H < 1$, while being suppressed (anti-persistence) for $0< H < 1/2$. $H=1/2$ corresponds to the classical Brownian motion. Similar to the previous case of the logistic map with varying control parameter $r$, we generate time series for varying $H \in (0, 1)$ with a step size of $\Delta H=0.05$ and present the resulting complexity--entropy planes in Fig.~\ref{fig:CEfBm}. Moreover, since fBm series in the persistent regime are nonstationary, we further transform all fBm realizations into stationary time series by employing first-order difference filtering, i.e., by considering the increments $x_{i+1}- x_i$. The transformed series are commonly referred to as fractional Gaussian noise (fGn), and the associated complexity--entropy planes are shown in Fig.~\ref{fig:CEfGn}. Notably, fGn retains the long-range correlations and Gaussian probability density function (PDF) of the underlying fBm process. For a discussion on practical concerns regarding time delay embedding of fBm and fGn processes, we refer to Ref.~\cite{Zou2015}. 
\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{CompEntropy_fBm.pdf}
\caption{\small{(Color online) Same as in Fig.~\ref{fig:CElogistic}, but for fractional Brownian motion with a Hurst exponent $H \in (0, 1)$ varying with a step size of $\Delta H=0.05$. Note that the normalized entropy is generally decreasing with increasing $H$ (red arrow).}  \label{fig:CEfBm}}
\end{figure*}

@@

{\color{red} For fBM, we can include another figure showing normalized complexity measures by the corresponding $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$ as Fig. \ref{fig:CHlogCmax}. As the embedding dimension $D$ is increased, the dots are more aligned to $\mathcal{C}_{max}$ and extend further to the left half of the complexity--entropy plane. 
\begin{figure*}
        \centering
        \includegraphics[width=2\columnwidth]{CompEntropyCNormalized_fBm.pdf}
\caption{(Color online) Three SCMs which are normalized by $\mathcal{C}_{max}$ and $\mathcal{C}_{min}$  at a given $\mathcal{H}$, for the fractional Brownian motion.  \label{fig:CHlogCmax}}
\end{figure*}
}

Comparing the results for both types of stochastic processes with those for the logistic map, a few important observations are made. First of all, the traditional pattern frequency based SCM values as originally reported by Rosso and co-workers \cite{RossoPRE2007,rossoPRL2007} are well reproduced as expected (Figs.~\ref{fig:CEfBm}(a, d) and \ref{fig:CEfGn}(a, d)). {\color{red}All three pairs of complexity measures lead to parabola-shaped curves as a function of the respective entropy. In addition, the newly proposed complexities of $\mathcal{C}_{W}$ and $\mathcal{C}_{E}$ have higher values, extending to the left plane of lower values for the entropy, which increases chances to discriminate chaos from stochastic processes in comparison to $\mathcal{C}_O$. }

Second, all complexity--entropy pairs of fBm clearly differ from the theoretical maximum complexity values, which does not change with increasing embedding dimension $D$. Especially for the new OPTN based SCMs, we find a partial overlap between the entropy values of fBm and the logistic map, while the associated SCM values are however clearly distinct except for the cases with the lowest entropy, which correspond to the most nonstationary situations ($H$ close to 1). {\color{red}Taking into account the effects of self loops in the transition matrix, the results of Fig. \ref{fig:CEfBm} do not change significantly although the SCM values extend slightly in the left half of the complexity--entropy plane as $H$ is increased, which are further shown in Fig. S1 of Supplementary Material (SM).} Accordingly, we suggest that combining information on the position in the complexity--entropy plane and on the stationarity could allow a clear distinction between chaotic and stochastic dynamics even in those extreme situations.

Notably, for the stationary fGn, all entropy measures are confined to very high entropy and low complexity values -- a range where no combinations have been found for the new transition frequency based SCMs for the logistic map. {\color{red}Furthermore, almost the same complexity--entropy plane has been obtained when self loops are included, which has been shown in Fig. S3 of SM. }

@@

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{logisticEntropy.pdf}
\caption{\small{Behavior of the different SCMs and associated entropy characteristics for the logistic map in dependence on the control parameter $r$ {\color{red}(embedding dimension $D = 6$).} (a) Bifurcation diagram, (b) Lyapunov exponent, (c) pattern frequency based (permutation) entropy $\mathcal{H}_O$ and (d) associated complexity measure $\mathcal{C}_O$; (e) pattern transition frequency based entropy $\mathcal{H}_W$ based on the globally normalized transition matrix $\mathbf{W}$ and (f) the corresponding complexity measure $\mathcal{C}_W$, and (g, h) entropy $\mathcal{H}_E$ and complexity measure $\mathcal{C}_E$ based on the node-wise out-link normalized transition matrix $\mathbf{W}$. Several major periodic windows have been highlighted by grey background shading. Vertical dotted lines indicate the cases summarized in Tab.~\ref{tableLog}. } \label{fig:bifurcation}}
\end{figure*}

@@

\begin{figure*}
	\centering 
	\includegraphics[width=2\columnwidth]{rosslerEntropy.pdf}
\caption{\small{Same as Fig. \ref{fig:bifurcation}, but for Poincar\'e sections of the R\"ossler system while varying the control parameter $b$ (see text for details {\color{red}and embedding dimension $D = 6$).} Panel (a) presents a bifurcation diagram based on the $x$ components of all points in the Poincar\'e sections. Panel (b) shows the largest Lyapunov exponent of the system in dependence of $b$.} \label{fig:bifurRossler}}
\end{figure*}

@@

Some open problems also common to other nonlinear time series analysis methods remain to be further addressed, for example, the dependence of reliable estimates of SCMs on the available time series length. {\color{red}For reliable estimations of traditional entropy $\mathcal{H}_O$ and SCM $\mathcal{C}_O$, the length $N > 5 D!$ has been largely suggested \cite{BandtPRL2002,AmigoPTRSA2014}. However, even longer length is required for the newly proposed SCMs based on transition frequencies because the transition matrix has the dimension $D!\times (D! - 1)$ if self loops are excluded. This should not be a big problem when working with time series from numerical models provided with the rapid increased computing power nowadays. } For time-discrete dynamical systems like chaotic maps, one may easily use longer data series, which however does not seem to improve the estimates very much due to the accumulation of effects due to a limited numerical precision (Fig.~\ref{fig:transient}). {\color{red} The dependence of the new SCMs on the length $N$ becomes crucial for experimental time series studies. One of more interesting problems is to understand the extra information of transition probability based complexity measures since linear correlations have been found for $\mathcal{C}_{E}$ and small entropy $H$, which are expected for a large number of patterns, i.e., $N = D!$ for high embedding $D$ or equivalently $N = D! \times (D!-1)$ transitions. } For continuous systems, together with the embedding parameters, the sampling frequency is another important characteristic which needs to be further explored. Moreover, the generalization of our approach from univariate to bivariate time series analysis shall be further discussed. We noticed that this aspect has remained a largely untouched subject for statistical complexity analysis so far, while it presents an interesting task to show the dependence of complexity on interaction. {\color{red} We may integrate some of the traditional network topological statistical measures of higher order into the SCMs, for instance, clustering coefficients or path lengths, which have been widely discussed to characterize the dynamical properties of the system generating the time series at hand \cite{ZouPR2018}.} We outline further work on the aforementioned topics as subjects of future research.

\section*{Acknowledgements}
Parts of this work have been financially supported by the National Natural Science Foundation of China (grant no. 11872182, 11835003, 11875132, 61973086), the Natural Science Foundation of Shanghai (grant no. 18ZR1411800). JZ and YZ further acknowledge support by the Shanghai Municipal Science and Technology Major Project (grant no. 2018SHZDZX01) and ZJ Lab. The authors thank J. B. Borges and O. A. Rosso for sharing algorithms for computing $\mathcal{C}_{min}$ and $\mathcal{C}_{max}$.
